Network based multifactorial modelling microRNA-target interaction.

### 2019-09-28 NAR_form_article.tex
article NAR form

---
title: Network based multifactorial modelling microRNA-target interaction.
author:
  - name: Selcen Ari
    affiliation: "1"
  - name: Alper Yilmaz
    affiliation: "1"
    correspond: "To whom correspondence should be addressed. Phone: +90 212 383 4627 Fax: +90 212 383 4625  Email: alyilmaz@yildiz.edu.tr"
    separator: "and"
address:
  - code: 1
    address: Department of Bioengineering, Yildiz Technical University, Istanbul, Turkey

abstract: |
  Competing endogenous RNA (ceRNA) regulations and crosstalk between various types of non-coding RNA in human is remarkable in means of miRNA regulation. Many studies have pointed out that an alteration in miRNA:target interaction can result in unexpected changes due to indirect and complex interactions. In this paper, we defined a new network-based model that handles miRNA:ceRNA interactions with expression values. Our model is able to handle miRNA interaction factors such as seed type, binding energy, if provided. Our approach is able to reveal that a perturbation in an element of network affects whole competing elements differently and cooperative efficiencies of miRNAs on common targets could be calculated. Our findings emphasized importance of miRNA:target ratios being crucial, as reported by previous studies. We have showed that the competing elements which have the same or close expression values may not be affected equally from the perturbation because of repression functionality depended on interaction factors of miRNA target pairs. We applied the model to real sample consisting of breast cancer gene and miRNA expression dataset and experimental miRNA:target interaction dataset all generated via high throughput sequencing methods. A gene over-expressed in tumor tissue, namely *SERPINE2*, is used as perturbing element. We have observed that change in expression level of single gene in miRNA:target network is sufficient to perturb regulations in whole network, due to unforeseen and unpredicted regulation which are only visible when considered in network context. Therefore, this model helps unveiling the crosstalk between elements in miRNA:target network where abundance of target and sponge effect are taken into account. The model is scalable and can be plugged in with emerging miRNA effectors such as circRNAs. The model is available as R package ceRNAnetsim https://github.com/selcenari/ceRNAnetsim.

bibliography: NAR-sample.bib
output: rticles::nar_article
header-includes: 
  \usepackage{float} \floatplacement{figure}{H} 
  \usepackage[colorlinks=true,
            linkcolor=blue,
            urlcolor=blue,
            citecolor=blue]{hyperref}
  \usepackage[T1]{fontenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# INTRODUCTION

MicroRNAs (miRNAs) are a family of short non-coding RNAs which are key regulator of gene expression through various post-transcriptional mechanisms. Although the mechanisms by which miRNA represses are not fully understood, miRNAs predominantly repress their targets. Repressive activities of miRNAs vary depending on many factors that are significant to miRNA:target interactions. These factors include miRNA:target binding energy, binding location in target sequence, base pairing types between miRNA and target, abundance of miRNAs and targets [@grimson_microrna_2007]. Binding energies of miRNA:target complexes vary based on nucleotide context and determine folding stability of complex [@cao_predicting_2012]. It has been demonstrated that the binding energy between miRNA and target indicates stability or affinity of complex [@helwak_mapping_2013] and does not directly determine repressive activity of miRNA [@cao_predicting_2012]. Early studies have argued that 2-8 nt sequence, seed, located in miRNA 5'end bind to specific sequence located in 3'UTR of its target [@bartel_micrornas_2004; @lewis_conserved_2005]. In recent studies, it has been shown that miRNAs can interact with targets via sequences located in regions such as 5’UTR or CDS [@hausser_analysis_2013; @helwak_mapping_2013; @moore_mirnatarget_2015]. These studies also showed that  binding location either indicate functionality of miRNA:target interaction or affect abundance of targets. It has been shown that miRNAs exhibit repressive activity via, 6-8 nt long sequence that is perfectly complementary with targets, seed region at the 5’ end of miRNAs [@bartel_micrornas:_2009; @grimson_microrna_2007]. On the other hand, some researchers have reported that seed sequence of miRNA can have mismatches or bulged/wobble nucleotides [@chi2012alternative] and may locate in region other than 5'end of miRNAs [@hafner_transcriptome-wide_2010; @helwak_mapping_2013]. On top of all these factors, abundance of miRNAs and targets and miRNA:target ratio in cells predominantly affect efficiency of miRNA:target interaction [@arvey_target_2010; @bosson_endogenous_2014; @denzler_assessing_2014].

As it is possible for miRNAs to suppress multiple targets, an individual mRNA molecule can also be targeted by multiple miRNAs. In that case, the targeted mRNAs exhibit competitor behavior, that is hypothesized as competing endogenous RNAs (ceRNAs) [@ala_integrated_2013; @cesana_deciphering_2013], against their miRNAs. Briefly, Ala et al. have explained the ceRNA hypothesis as disturbance of the other target when one of the targets on a steady-state system that included one miRNA and two target was perturbed with expression change [@ala_integrated_2013]. Regarding interaction between miRNAs and their target in a cell, explaining and predicting results of an individual perturbation is difficult due to complexity of interactions. Various computational and experimental studies have tackled the problem of unraveling ceRNA:miRNA interactions.

@@

It has been observed that when abundance of one of the targets of miR-122 was increased, the other targets' expression also slightly increased as a result of decreasing repressive activity of miR-122 on them [@denzler_assessing_2014]. Bosson et al. have developed a mathematical model for changes on total target pool concentration after grouping targets according to affinity and demonstrated that miRNA activity correlated with affinity between miRNA and target [@bosson_endogenous_2014]. Cooperative efficiency of miRNAs as well as competitor behaviors of targets were also studied and it has been demonstrated to be crucial for regulating available mRNA levels of targets [@denzler_impact_2016]. MiRNA:target interactions have been modeled as stoichiometric and catalytic mechanisms and Figliuzzi et al. have recommended handling models in network context [@figliuzzi_micrornas_2013]. The model that can explain miRNA target interaction through topological features has been applied at bipartite network by Nitzan et al. [@nitzan_interactions_2014]. Robinson and Henderson applied the model that handles miRNA:target direct and indirect interactions via common targeting miRNA of genes and target of miRNAs, at bipartite network. It has been demonstrated that all miRNAs and targets in the network can interact with each other through common miRNAs and genes, without interaction between the same type of nodes [@robinson_modelling_2018].Associated genes that are targets of the same miRNAs have been found with help of correlation of gene expression changes in recent algorithm [@markus_list_sponge_2017]. List et al. have specified that their approach can be useful for ceRNA studies and published their approach as an R package.

# MATERIALS AND METHODS

## Construction of miRNA:target network
miRNA and target pairs per line should be provided as edge list to construct the network. At each line minimum required information is expression levels of miRNA and the target. If available, additional data about factors effecting binding or efficiency of miRNA can be provided as separate columns. After construction of the network, amount of miRNA per target is calculated and kept as edge data. Simply, a target will sequester miRNA proportional to its ratio amount among other targets. If additional criteria effecting the binding of miRNA to its target is provided, distribution of miRNA will be calculated accordingly. Target can be mRNA or any other ceRNA (circRNA, ncRNA, etc.) thus, throughout the manuscript terms target, gene and ceRNA are used interchangeably.

## Triggering perturbation and subsequent calculations
Initially, the network is assumed in steady-state (Figure \ref{fig:fig1}A) condition and needs least one trigger for initiating calculations. The trigger can be a change in expression level of one or more genes (Figure \ref{fig:fig1}B). After a trigger, the network undergoes iterative cycle of calculations at each of which distribution of miRNA in local neighborhood is recalculated (Figure \ref{fig:fig1}C). Based on new miRNA distribution, expression level of each node (i.e. ceRNA) is updated. This update results in change of expression value of common target (G4) in system. In this case, the common target acts as trigger for the other group (the targets interacted with M2 miRNA) in the network. Due to common targeted elements, the change in one neighborhood spreads to other neighborhoods (Figure \ref{fig:fig1}D), consequently have potential to affect whole network due to "ripple effect".

@@

The repression efficiency of a miRNA on the individual target ($Eff_{gi}$) is calculated according to equation \eqref{eq:1}; where miRNA expression ($C_m$) in local neighborhood is distributed among targets using individual gene expression levels ($C_{gi}$). For the genes targeted by multiple miRNAs, cooperative activity of miRNAs on a target gene, $R$, is calculated by summing repression activity of each miRNA Equation \eqref{eq:2}.
\begin{equation*} 
    Eff_{gi}= C_m \times C_{gi}/\sum_{1}^{i} C_{gi} \tag{1}\label{eq:1}
\end{equation*}

@@

## Multifactorial calculations in miRNA:target network
Interactions between miRNAs and their targets can be affected from various factors. So, our model integrates multiple factors when calculating overall miRNA activity. We classified factors into two categories. Binding factors determine interaction between miRNA and target and they alter amount of miRNA sequestered to target. Efficiency factors dictate degradation efficiency of miRNA on its target and define amount of target to be degraded. In other words, binding factors exert their influence before or during binding, efficiency factors exert their influence after binding. In the literature, binding free energy [@cao_predicting_2012 ; @helwak_mapping_2013] and seed type [@werfel_preferential_2017] in miRNA:target interactions are described as factors effecting binding affinity. Also, binding region (e.g, 5'UTR, 3'UTR, CDS) was shown to drastically affect miRNA degradation efficiency [@hausser_analysis_2013; @helwak_mapping_2013]. Since binding or degradation efficiency from literature have various and broad ranges, both binding and efficiency factors are normalized to their maximum values and scaled to [0,1] interval. The normalized values of factors are used to determine binding activity and miRNA efficiency on targets (Figure \ref{fig:fig2}). Binding affinities (activity, $Eff$) of miRNAs on each individual gene are calculated as shown in equation \eqref{eq:3}; $C_m$, miRNA expression in the group; $C_{gi}$, individual gene expression; gi, individual gene (Figure \ref{fig:fig2}C).

@@

Not all miRNA:target binding events result in degradation of target. The degradation of target by bound miRNA depends on efficiency factors such as binding region. Exact repression efficiency of miRNA is calculated according to equation \eqref{eq:4} (Figure \ref{fig:fig2}D); $RE^\prime_{gi}$, normalized values of region efficiency coefficient between miRNA and gene. The cooperative repression activity of miRNAs to their common targets is figured out as shown in Figure \ref{fig:fig2}E.

## Breast cancer patient dataset

We have applied our model in a real dataset for which experimental measurements of various factors were available. Expression levels of miRNA and genes in tumor and normal tissue of single patient are retrieved from TCGA Research Network https://www.cancer.gov/tcga. High-throughput experimental datasets which are provided miRNA:gene target pairs with interaction factors [@helwak_mapping_2013; @moore_mirnatarget_2015]. We have combined miRNA and gene expression datasets via miRNA:target gene dataset retrieved from published datasets [@helwak_mapping_2013; @moore_mirnatarget_2015]. Detailed description of network construction and its code is available in Supplementary Materials and Methods. SERPINE2, one of the most effective nodes, has been used as trigger in ceRNA network simulation in breast cancer patient network. Relevant steps and codes are available at  Supplementary Materials and Methods.

[#TODO]: <> "about formulas, 2) in formulas eq:1 and eq:2, C_m or Eff_gi is not intuitive, is there consensus terms about effect or nth gene?"

# RESULTS

## Network with miRNA:ceRNA expression level
We have developed a network-based approach to assess effects of expression level changes in competitive ceRNA regulation. The basic mode of miRNA repression activity has been based on miRNA and target abundance in various researches [@arvey_target_2010; @denzler_assessing_2014]. Our approach can effortlessly calculate effects of expression changes when abundance levels of miRNAs and targets is only available factor. In sample network given in Figure (\ref{fig:fig1}), after an increase in expression level of a gene (G2), expression values of other genes also changed due to redistribution of miRNA among its targets. Previous studies have shown that if a gene abundance increases in ceRNA system, expression levels of genes targeted by shared miRNA are also affected [@lai_understanding_2016; @salmena_cerna_2011; @tay_multilayered_2014]. It was observed that expression levels of primary neighborhoods which do not interact with another miRNA of the trigger gene change in relation to their expression. However, changes in expression levels of G4 and G1 differ although they have same initial conditions. This is due to fact that G4 is targeted by two miRNAs, M1 and M2 (Figure \ref{fig:fig1}D).

Genes targeted by multiple miRNAs act as a trigger for adjacent local neighborhood of targeting miRNAs, causing changes in expression levels of genes outside the local neighborhood of original trigger gene. Therefore, primary expression level change in gene (G2) causes changes in other group of genes (G5 and G6) even though original trigger gene (G2) and genes in other group are not targeted by common miRNA. In addition, as shown by ceRNA hypothesis model of Ala et al., after the increase of gene expression level of G2, the miRNA that is found in the same group (M1) tended to be less repressive on its remaining targets (G1, G3 and G4). It's important to note that the changes in gene expression levels will have more pronounced effect if miRNA:target ratio is high, i.e., more miRNA available per target, which was reported in previous findings [@arvey_target_2010; @bosson_endogenous_2014; @denzler_assessing_2014].

@@

## ceRNA:Target networks based on interaction factors
Earlier studies reported that miRNA regulatory interactions are affected by different parameters. For example, Xu et al. have investigated the importance of seed pairing type between miRNAs and their targets and target site location by using proteomics dataset [@xu_characterization_2014]. They have proposed that the features of binding between miRNA and target can be critical for miRNA efficiency. In addition, binding energy between miRNA and targets is a significant determinant for miRNA efficiency and it has been reported that strength of miRNA:target interactions is depended on binding energy of complexes [@breda_quantifying_2015]. Similarly, another study revealed that affinity is correlated with seed pairing of miRNA:target pairs and suggested that affinity is correlated with number of canonical seed base pairing [@bosson_endogenous_2014]. Therefore, we have integrated aforementioned interaction parameters that could be useful for predicting miRNA repression activity more accurately. The sample dataset used in Figure \ref{fig:fig1} is recalculated with additional factors Table \ref{tab:one} in effect with same trigger, two fold increase in Gene2 (Figure \ref{fig:fig3}). When the factors were taken into account in the system, miRNA efficiencies varied as shown in Figure \ref{fig:fig3}A. Although the miRNA:target expression ratios in steady-state were same in comparison with the sample dataset without factors, efficiency of binding and repression have changed.

[#TODO]: <> "figure1, 2 ve 3 uzerinden farkli dagilim ve degradasyon açıklaması buraya. G2 uzerinden olmasa iyi olur"

@@

In regulation of miRNA target sample system (Figure \ref{fig:fig3}), miRNA (M1) repressive efficiency on the primary triggering gene (G2) is low in steady-state. So it has been observed that the regulatory activities of miRNA on the targets are weak after the increase of miRNA target G2. When model is triggered with two fold increase in expression level of common target (G4), more prominent changes were observed in gene expression levels in network compared to changes caused by G2 increase. Furthermore, change in expression level of target gene that has strong miRNA repression efficiency resulted in evident perturbation in network. On the other hand, it was observed that Gene2 was weakly affected from change of Gene4 expression because of its weak interaction factors. Since perturbation efficiency of each gene is different, we have developed a function which screens each gene in the network for their perturbation efficiencies. *(see details in Supplementary Materials and Methods*. On the other hand, when we applied the method on the minimal dataset, Gene4 has been found to be the most efficient element in terms of number of perturbed elements and miRNA2 (M2) has been found to be causing the highest mean in expression changes.

## Simulation of large and realistic network

Apart from tools predicting miRNA:target pairs, there are databases that provide  experimentally validated miRNA:target interaction datasets. The reliability in these datasets range from weak (e.g, western blot) to strong (e.g, CLiP, CLASH). High-throughput methods ensure handling of AGO (Argounate Protein) interacted miRNAs (CLiP) and their targets as well (CLEAR-CLiP, CLASH). We constructed a large ceRNA network utilizing high-throughput datasets containing chimeric reading of miRNA:target pairs [@helwak_mapping_2013; @moore_mirnatarget_2015] which expose exact complementary seed regions.

Although these experimental sources include exact binding sites, they do not imply that each miRNA binding event is functional [@liu2019prediction]. Seed type, energy, binding region affects functionality of miRNA drastically. In order to construct realistic network, we incorporated such factors in the network. Energy  values in miRNA:target pairs represented by high-throughput studies [@helwak_mapping_2013; @moore_mirnatarget_2015], seed type [@grimson_microrna_2007; @bartel_micrornas:_2009; @betel2010comprehensive] and binding region effect [@hausser_analysis_2013; @helwak_mapping_2013] data were collected and integrated  into network.

The network was finalized by integrating gene and miRNA expression levels from TCGA determined by RNA-Seq. Our  model can successfully simulate perturbations in such large network despite complex behaviors and struggle to reach steady-state *(see details in Supplementary Materials and Methods)*. Simulations show that change in expression level of single gene has potential to affect whole network, perturbing almost all nodes. These observations are in accordance with competing endogenous RNA hypothesis where genes targeted with many common miRNAs subsequently transmit perturbation to neighboring groups.

We have observed that the results of simulation with SERPINE2 gene, a serine protease inhibitor, one of the overexpressed genes of breast cancer [@yang2018expression; @candia2006protease] is not convenient with tumor tissue expression values, when we compared the tumor gene expression dataset retrieved from TCGA and simulation results obtained via our approach. Simulation results were not expected to be consistent with tumor tissue expression values because large number of abnormal regulations were observed in these tissues. So, it might not said that a single gene is responsible for entire regulations in cancer cells. Besides, the all other factors such as up/down regulation of miRNAs or other genes were ignored. Therefore, it would be more useful to test our approach with an ideal dataset that includes entire informations of gene expression values before and after the regulation of a known gene at existing tissue, and miRNA expression values of this tissue at initial conditions.

# DISCUSSION

>>>>

\documentclass[a4,center,fleqn]{NAR}

% Enter dates of publication
\copyrightyear{2008}
\pubdate{31 July 2009}
\pubyear{2009}
\jvolume{37}
\jissue{12}

%\articlesubtype{This is the article type (optional)}

\usepackage{float}

\floatplacement{figure}{H}

\usepackage[colorlinks=true, linkcolor=blue, urlcolor=blue, citecolor=blue]{hyperref} \usepackage[T1]{fontenc}

\begin{document}

\title{Network based multifactorial modelling microRNA-target interaction.}

\author{
    Selcen Ari\,$^{1}$
   and Alper Yilmaz\,$^{1}$ \footnote{To whom correspondence should be addressed. Phone: +90 212 383 4627 Fax:
+90 212 383 4625 Email:
\href{mailto:alyilmaz@yildiz.edu.tr}{\nolinkurl{alyilmaz@yildiz.edu.tr}}} 
}

\address{
   $^{1}$ Department of Bioengineering, Yildiz Technical University, Istanbul,
Turkey 
}

\history{%
Received January 1, 2009;
Revised February 1, 2009;
Accepted March 1, 2009}

\maketitle

\begin{abstract}
Competing endogenous RNA (ceRNA) regulations and crosstalk between
various types of non-coding RNA in human is remarkable in means of miRNA
regulation. Many studies have pointed out that an alteration in
miRNA:target interaction can result in unexpected changes due to
indirect and complex interactions. In this paper, we defined a new
network-based model that handles miRNA:ceRNA interactions with
expression values. Our model is able to handle miRNA interaction factors
such as seed type, binding energy, if provided. Our approach is able to
reveal that a perturbation in an element of network affects whole
competing elements differently and cooperative efficiencies of miRNAs on
common targets could be calculated. Our findings emphasized importance
of miRNA:target ratios being crucial, as reported by previous studies.
We have showed that the competing elements which have the same or close
expression values may not be affected equally from the perturbation
because of repression functionality depended on interaction factors of
miRNA target pairs. We applied the model to real sample consisting of
breast cancer gene and miRNA expression dataset and experimental
miRNA:target interaction dataset all generated via high throughput
sequencing methods. A gene over-expressed in tumor tissue, namely
\emph{SERPINE2}, is used as perturbing element. We have observed that
change in expression level of single gene in miRNA:target network is
sufficient to perturb regulations in whole network, due to unforeseen
and unpredicted regulation which are only visible when considered in
network context. Therefore, this model helps unveiling the crosstalk
between elements in miRNA:target network where abundance of target and
sponge effect are taken into account. The model is scalable and can be
plugged in with emerging miRNA effectors such as circRNAs. The model is
available as R package ceRNAnetsim
\url{https://github.com/selcenari/ceRNAnetsim}.
\end{abstract}

\section{INTRODUCTION}

MicroRNAs (miRNAs) are a family of short non-coding RNAs which are key
regulator of gene expression through various post-transcriptional
mechanisms. Although the mechanisms by which miRNA represses are not
fully understood, miRNAs predominantly repress their targets. Repressive
activities of miRNAs vary depending on many factors that are significant
to miRNA:target interactions. These factors include miRNA:target binding
energy, binding location in target sequence, base pairing types between
miRNA and target, abundance of miRNAs and targets
\citep{grimson_microrna_2007}. Binding energies of miRNA:target
complexes vary based on nucleotide context and determine folding
stability of complex \citep{cao_predicting_2012}. It has been
demonstrated that the binding energy between miRNA and target indicates
stability or affinity of complex \citep{helwak_mapping_2013} and does
not directly determine repressive activity of miRNA
\citep{cao_predicting_2012}. Early studies have argued that 2-8 nt
sequence, seed, located in miRNA 5'end bind to specific sequence located
in 3'UTR of its target
\citep{bartel_micrornas_2004, lewis_conserved_2005}. In recent studies,
it has been shown that miRNAs can interact with targets via sequences
located in regions such as 5'UTR or CDS
\citep{hausser_analysis_2013, helwak_mapping_2013, moore_mirnatarget_2015}.
These studies also showed that binding location either indicate
functionality of miRNA:target interaction or affect abundance of
targets. It has been shown that miRNAs exhibit repressive activity via,
6-8 nt long sequence that is perfectly complementary with targets, seed
region at the 5' end of miRNAs
\citep{bartel_micrornas:_2009, grimson_microrna_2007}. On the other
hand, some researchers have reported that seed sequence of miRNA can
have mismatches or bulged/wobble nucleotides \citep{chi2012alternative}
and may locate in region other than 5'end of miRNAs
\citep{hafner_transcriptome-wide_2010, helwak_mapping_2013}. On top of
all these factors, abundance of miRNAs and targets and miRNA:target
ratio in cells predominantly affect efficiency of miRNA:target
interaction
\citep{arvey_target_2010, bosson_endogenous_2014, denzler_assessing_2014}.

As it is possible for miRNAs to suppress multiple targets, an individual
mRNA molecule can also be targeted by multiple miRNAs. In that case, the
targeted mRNAs exhibit competitor behavior, that is hypothesized as
competing endogenous RNAs (ceRNAs)
\citep{ala_integrated_2013, cesana_deciphering_2013}, against their
miRNAs. Briefly, Ala et al.~have explained the ceRNA hypothesis as
disturbance of the other target when one of the targets on a
steady-state system that included one miRNA and two target was perturbed
with expression change \citep{ala_integrated_2013}. Regarding
interaction between miRNAs and their target in a cell, explaining and
predicting results of an individual perturbation is difficult due to
complexity of interactions. Various computational and experimental
studies have tackled the problem of unraveling ceRNA:miRNA interactions.

@@

It has been observed that when abundance of one of the targets of
miR-122 was increased, the other targets' expression also slightly
increased as a result of decreasing repressive activity of miR-122 on
them \citep{denzler_assessing_2014}. Bosson et al.~have developed a
mathematical model for changes on total target pool concentration after
grouping targets according to affinity and demonstrated that miRNA
activity correlated with affinity between miRNA and target
\citep{bosson_endogenous_2014}. Cooperative efficiency of miRNAs as well
as competitor behaviors of targets were also studied and it has been
demonstrated to be crucial for regulating available mRNA levels of
targets \citep{denzler_impact_2016}. MiRNA:target interactions have been
modeled as stoichiometric and catalytic mechanisms and Figliuzzi et
al.~have recommended handling models in network context
\citep{figliuzzi_micrornas_2013}. The model that can explain miRNA
target interaction through topological features has been applied at
bipartite network by Nitzan et al.~\citep{nitzan_interactions_2014}.
Robinson and Henderson applied the model that handles miRNA:target
direct and indirect interactions via common targeting miRNA of genes and
target of miRNAs, at bipartite network. It has been demonstrated that
all miRNAs and targets in the network can interact with each other
through common miRNAs and genes, without interaction between the same
type of nodes \citep{robinson_modelling_2018}.Associated genes that are
targets of the same miRNAs have been found with help of correlation of
gene expression changes in recent algorithm
\citep{markus_list_sponge_2017}. List et al.~have specified that their
approach can be useful for ceRNA studies and published their approach as
an R package.

\section{MATERIALS AND METHODS}

\subsection{Construction of miRNA:target network}

miRNA and target pairs per line should be provided as edge list to
construct the network. At each line minimum required information is
expression levels of miRNA and the target. If available, additional data
about factors effecting binding or efficiency of miRNA can be provided
as separate columns. After construction of the network, amount of miRNA
per target is calculated and kept as edge data. Simply, a target will
sequester miRNA proportional to its ratio amount among other targets. If
additional criteria effecting the binding of miRNA to its target is
provided, distribution of miRNA will be calculated accordingly. Target
can be mRNA or any other ceRNA (circRNA, ncRNA, etc.) thus, throughout
the manuscript terms target, gene and ceRNA are used interchangeably.

\subsection{Triggering perturbation and subsequent calculations}

Initially, the network is assumed in steady-state (Figure
\ref{fig:fig1}A) condition and needs least one trigger for initiating
calculations. The trigger can be a change in expression level of one or
more genes (Figure \ref{fig:fig1}B). After a trigger, the network
undergoes iterative cycle of calculations at each of which distribution
of miRNA in local neighborhood is recalculated (Figure \ref{fig:fig1}C).
Based on new miRNA distribution, expression level of each node
(i.e.~ceRNA) is updated. This update results in change of expression
value of common target (G4) in system. In this case, the common target
acts as trigger for the other group (the targets interacted with M2
miRNA) in the network. Due to common targeted elements, the change in
one neighborhood spreads to other neighborhoods (Figure
\ref{fig:fig1}D), consequently have potential to affect whole network
due to ``ripple effect''.

@@

The repression efficiency of a miRNA on the individual target
(\(Eff_{gi}\)) is calculated according to equation \eqref{eq:1}; where
miRNA expression (\(C_m\)) in local neighborhood is distributed among
targets using individual gene expression levels (\(C_{gi}\)). For the
genes targeted by multiple miRNAs, cooperative activity of miRNAs on a
target gene, \(R\), is calculated by summing repression activity of each
miRNA Equation \eqref{eq:2}. \begin{equation*} 
    Eff_{gi}= C_m \times C_{gi}/\sum_{1}^{i} C_{gi} \tag{1}\label{eq:1}
\end{equation*}

@@

\subsection{Multifactorial calculations in miRNA:target network}

Interactions between miRNAs and their targets can be affected from
various factors. So, our model integrates multiple factors when
calculating overall miRNA activity. We classified factors into two
categories. Binding factors determine interaction between miRNA and
target and they alter amount of miRNA sequestered to target. Efficiency
factors dictate degradation efficiency of miRNA on its target and define
amount of target to be degraded. In other words, binding factors exert
their influence before or during binding, efficiency factors exert their
influence after binding. In the literature, binding free energy
\citetext{\citealp[
]{cao_predicting_2012}; \citealp{helwak_mapping_2013}} and seed type
\citep{werfel_preferential_2017} in miRNA:target interactions are
described as factors effecting binding affinity. Also, binding region
(e.g, 5'UTR, 3'UTR, CDS) was shown to drastically affect miRNA
degradation efficiency
\citep{hausser_analysis_2013, helwak_mapping_2013}. Since binding or
degradation efficiency from literature have various and broad ranges,
both binding and efficiency factors are normalized to their maximum
values and scaled to {[}0,1{]} interval. The normalized values of
factors are used to determine binding activity and miRNA efficiency on
targets (Figure \ref{fig:fig2}). Binding affinities (activity, \(Eff\))
of miRNAs on each individual gene are calculated as shown in equation
\eqref{eq:3}; \(C_m\), miRNA expression in the group; \(C_{gi}\),
individual gene expression; gi, individual gene (Figure
\ref{fig:fig2}C).

@@

Not all miRNA:target binding events result in degradation of target. The
degradation of target by bound miRNA depends on efficiency factors such
as binding region. Exact repression efficiency of miRNA is calculated
according to equation \eqref{eq:4} (Figure \ref{fig:fig2}D);
\(RE^\prime_{gi}\), normalized values of region efficiency coefficient
between miRNA and gene. The cooperative repression activity of miRNAs to
their common targets is figured out as shown in Figure \ref{fig:fig2}E.

\subsection{Breast cancer patient dataset}

We have applied our model in a real dataset for which experimental
measurements of various factors were available. Expression levels of
miRNA and genes in tumor and normal tissue of single patient are
retrieved from TCGA Research Network \url{https://www.cancer.gov/tcga}.
High-throughput experimental datasets which are provided miRNA:gene
target pairs with interaction factors
\citep{helwak_mapping_2013, moore_mirnatarget_2015}. We have combined
miRNA and gene expression datasets via miRNA:target gene dataset
retrieved from published datasets
\citep{helwak_mapping_2013, moore_mirnatarget_2015}. Detailed
description of network construction and its code is available in
Supplementary Materials and Methods. SERPINE2, one of the most effective
nodes, has been used as trigger in ceRNA network simulation in breast
cancer patient network. Relevant steps and codes are available at
Supplementary Materials and Methods.

\section{RESULTS}

\subsection{Network with miRNA:ceRNA expression level}

We have developed a network-based approach to assess effects of
expression level changes in competitive ceRNA regulation. The basic mode
of miRNA repression activity has been based on miRNA and target
abundance in various researches
\citep{arvey_target_2010, denzler_assessing_2014}. Our approach can
effortlessly calculate effects of expression changes when abundance
levels of miRNAs and targets is only available factor. In sample network
given in Figure (\ref{fig:fig1}), after an increase in expression level
of a gene (G2), expression values of other genes also changed due to
redistribution of miRNA among its targets. Previous studies have shown
that if a gene abundance increases in ceRNA system, expression levels of
genes targeted by shared miRNA are also affected
\citep{lai_understanding_2016, salmena_cerna_2011, tay_multilayered_2014}.
It was observed that expression levels of primary neighborhoods which do
not interact with another miRNA of the trigger gene change in relation
to their expression. However, changes in expression levels of G4 and G1
differ although they have same initial conditions. This is due to fact
that G4 is targeted by two miRNAs, M1 and M2 (Figure \ref{fig:fig1}D).

Genes targeted by multiple miRNAs act as a trigger for adjacent local
neighborhood of targeting miRNAs, causing changes in expression levels
of genes outside the local neighborhood of original trigger gene.
Therefore, primary expression level change in gene (G2) causes changes
in other group of genes (G5 and G6) even though original trigger gene
(G2) and genes in other group are not targeted by common miRNA. In
addition, as shown by ceRNA hypothesis model of Ala et al., after the
increase of gene expression level of G2, the miRNA that is found in the
same group (M1) tended to be less repressive on its remaining targets
(G1, G3 and G4). It's important to note that the changes in gene
expression levels will have more pronounced effect if miRNA:target ratio
is high, i.e., more miRNA available per target, which was reported in
previous findings
\citep{arvey_target_2010, bosson_endogenous_2014, denzler_assessing_2014}.

@@

\subsection{ceRNA:Target networks based on interaction factors}

Earlier studies reported that miRNA regulatory interactions are affected
by different parameters. For example, Xu et al.~have investigated the
importance of seed pairing type between miRNAs and their targets and
target site location by using proteomics dataset
\citep{xu_characterization_2014}. They have proposed that the features
of binding between miRNA and target can be critical for miRNA
efficiency. In addition, binding energy between miRNA and targets is a
significant determinant for miRNA efficiency and it has been reported
that strength of miRNA:target interactions is depended on binding energy
of complexes \citep{breda_quantifying_2015}. Similarly, another study
revealed that affinity is correlated with seed pairing of miRNA:target
pairs and suggested that affinity is correlated with number of canonical
seed base pairing \citep{bosson_endogenous_2014}. Therefore, we have
integrated aforementioned interaction parameters that could be useful
for predicting miRNA repression activity more accurately. The sample
dataset used in Figure \ref{fig:fig1} is recalculated with additional
factors Table \ref{tab:one} in effect with same trigger, two fold
increase in Gene2 (Figure \ref{fig:fig3}). When the factors were taken
into account in the system, miRNA efficiencies varied as shown in Figure
\ref{fig:fig3}A. Although the miRNA:target expression ratios in
steady-state were same in comparison with the sample dataset without
factors, efficiency of binding and repression have changed.

@@

In regulation of miRNA target sample system (Figure \ref{fig:fig3}),
miRNA (M1) repressive efficiency on the primary triggering gene (G2) is
low in steady-state. So it has been observed that the regulatory
activities of miRNA on the targets are weak after the increase of miRNA
target G2. When model is triggered with two fold increase in expression
level of common target (G4), more prominent changes were observed in
gene expression levels in network compared to changes caused by G2
increase. Furthermore, change in expression level of target gene that
has strong miRNA repression efficiency resulted in evident perturbation
in network. On the other hand, it was observed that Gene2 was weakly
affected from change of Gene4 expression because of its weak interaction
factors. Since perturbation efficiency of each gene is different, we
have developed a function which screens each gene in the network for
their perturbation efficiencies. \emph{(see details in Supplementary
Materials and Methods}. On the other hand, when we applied the method on
the minimal dataset, Gene4 has been found to be the most efficient
element in terms of number of perturbed elements and miRNA2 (M2) has
been found to be causing the highest mean in expression changes.

\subsection{Simulation of large and realistic network}

Apart from tools predicting miRNA:target pairs, there are databases that
provide experimentally validated miRNA:target interaction datasets. The
reliability in these datasets range from weak (e.g, western blot) to
strong (e.g, CLiP, CLASH). High-throughput methods ensure handling of
AGO (Argounate Protein) interacted miRNAs (CLiP) and their targets as
well (CLEAR-CLiP, CLASH). We constructed a large ceRNA network utilizing
high-throughput datasets containing chimeric reading of miRNA:target
pairs \citep{helwak_mapping_2013, moore_mirnatarget_2015} which expose
exact complementary seed regions.

Although these experimental sources include exact binding sites, they do
not imply that each miRNA binding event is functional
\citep{liu2019prediction}. Seed type, energy, binding region affects
functionality of miRNA drastically. In order to construct realistic
network, we incorporated such factors in the network. Energy values in
miRNA:target pairs represented by high-throughput studies
\citep{helwak_mapping_2013, moore_mirnatarget_2015}, seed type
\citep{grimson_microrna_2007, bartel_micrornas:_2009, betel2010comprehensive}
and binding region effect
\citep{hausser_analysis_2013, helwak_mapping_2013} data were collected
and integrated into network.

The network was finalized by integrating gene and miRNA expression
levels from TCGA determined by RNA-Seq. Our model can successfully
simulate perturbations in such large network despite complex behaviors
and struggle to reach steady-state \emph{(see details in Supplementary
Materials and Methods)}. Simulations show that change in expression
level of single gene has potential to affect whole network, perturbing
almost all nodes. These observations are in accordance with competing
endogenous RNA hypothesis where genes targeted with many common miRNAs
subsequently transmit perturbation to neighboring groups.

We have observed that the results of simulation with SERPINE2 gene, a
serine protease inhibitor, one of the overexpressed genes of breast
cancer \citep{yang2018expression, candia2006protease} is not convenient
with tumor tissue expression values, when we compared the tumor gene
expression dataset retrieved from TCGA and simulation results obtained
via our approach. Simulation results were not expected to be consistent
with tumor tissue expression values because large number of abnormal
regulations were observed in these tissues. So, it might not said that a
single gene is responsible for entire regulations in cancer cells.
Besides, the all other factors such as up/down regulation of miRNAs or
other genes were ignored. Therefore, it would be more useful to test our
approach with an ideal dataset that includes entire informations of gene
expression values before and after the regulation of a known gene at
existing tissue, and miRNA expression values of this tissue at initial
conditions.

\section{DISCUSSION}
